---
title: "fastRanges: Scalable Genomic Interval Operations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fastRanges: Scalable Genomic Interval Operations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r vignette-options, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

In Bioconductor, genomic intervals are represented by `GRanges` objects
(`seqnames`, genomic coordinates, and strand). Core interval overlap operations
are provided by `findOverlaps()` from `IRanges` and `GenomicRanges`.

`fastRanges` provides deterministic, multithreaded overlap and summary
operations for `IRanges` and `GRanges`, with APIs designed to stay close to
Bioconductor semantics while improving throughput.

```{r setup}
library(fastRanges)
library(GenomicRanges)

data("fast_ranges_example", package = "fastRanges")
query <- fast_ranges_example$query
subject <- fast_ranges_example$subject
```

## Overlap and Count

```{r overlaps}
hits <- fast_find_overlaps(query, subject, threads = 2)
counts <- fast_count_overlaps(query, subject, threads = 2)
any_hits <- fast_overlaps_any(query, subject, threads = 2)

length(hits)
counts
any_hits
```

## Bioconductor Compatibility

`fastRanges` keeps the main overlap argument grammar aligned with Bioconductor:

- `query`, `subject`, `max_gap`, `min_overlap`, `type`, `ignore_strand`

Additional package-specific controls:

- `threads` for explicit parallelism
- `deterministic` for reproducible ordering of hit pairs

Typical return classes:

```{r return-types}
ret_classes <- c(
  find = class(fast_find_overlaps(query, subject, threads = 1))[1],
  count = class(fast_count_overlaps(query, subject, threads = 1))[1],
  any = class(fast_overlaps_any(query, subject, threads = 1))[1],
  nearest = class(fast_nearest(query, subject))[1]
)

ret_classes
```

## Reusable Subject Index

For repeated queries, build a subject index once and reuse it.

```{r index}
subject_index <- fast_build_index(subject)
hits_from_index <- fast_find_overlaps(query, subject_index, threads = 2)
length(hits_from_index)
```

The indexed and direct queries return equivalent overlap mappings:

```{r index-equivalence}
direct_hits <- fast_find_overlaps(query, subject, threads = 2)

identical(
  cbind(S4Vectors::queryHits(hits_from_index), S4Vectors::subjectHits(hits_from_index)),
  cbind(S4Vectors::queryHits(direct_hits), S4Vectors::subjectHits(direct_hits))
)
```

## Overlap Joins

`fast_overlap_join()` returns an overlap pair table with prefixed query and subject metadata columns.

```{r joins}
joined <- fast_overlap_join(query, subject, join = "inner", threads = 2)
head(joined)
```

## Nearest Queries

```{r nearest}
nearest_tbl <- fast_nearest(query, subject)
head(as.data.frame(nearest_tbl))
```

## Extended Overlap Helpers

```{r derived-overlap-helpers}
self_hits <- fast_self_overlaps(query, threads = 2)
clusters <- fast_cluster_overlaps(query)
window_counts <- fast_window_count_overlaps(query, subject, window_width = 1000L, step_width = 500L, threads = 2)

c(self_hits = length(self_hits), clusters = length(unique(clusters)))
head(window_counts)
```

## Grouped Counts and Aggregation

```{r grouped-aggregation}
set.seed(1)
S4Vectors::mcols(subject)$group <- sample(c("promoter", "enhancer"), length(subject), replace = TRUE)
S4Vectors::mcols(subject)$score <- seq_len(length(subject))

by_group <- fast_count_overlaps_by_group(query, subject, group_col = "group", threads = 2)
sum_score <- fast_overlap_aggregate(query, subject, value_col = "score", fun = "sum", threads = 2)

dim(by_group)
head(sum_score)
```

## Range Algebra and Coverage

```{r range-coverage}
query_reduced <- fast_reduce(query)
query_disjoint <- fast_disjoin(query)
query_cov <- fast_coverage(query)
query_tiles <- fast_tile_coverage(query, tile_width = 1000L, step_width = 1000L)

c(reduced = length(query_reduced), disjoint = length(query_disjoint), coverage_seq = length(query_cov))
head(query_tiles)
```

## Iterator + Index Lifecycle

```{r iterator-index}
subject_index <- fast_build_index(subject)
tmp_index <- tempfile(fileext = ".rds")
fast_save_index(subject_index, tmp_index)
loaded_index <- fast_load_index(tmp_index)
fast_index_stats(loaded_index)

iter <- fast_find_overlaps_iter(query, loaded_index, chunk_size = 100L, threads = 2)
iter_hits <- fast_iter_collect(iter)
length(iter_hits)
unlink(tmp_index)
```

## Quick Benchmark

```{r benchmark}
# Small example benchmark for illustration only.
# For publication-scale runs, use inst/benchmarks/benchmark_bioc.qmd.
base_time <- system.time({
  h_base <- GenomicRanges::findOverlaps(query, subject, ignore.strand = FALSE)
})

subject_index <- fast_build_index(subject)
fast_time <- system.time({
  h_fast <- fast_find_overlaps(query, subject_index, ignore_strand = FALSE, threads = 2)
})

rbind(
  genomic_ranges = base_time[1:3],
  fast_ranges = fast_time[1:3]
)

c(base_hits = length(h_base), fast_hits = length(h_fast))
```

## External Example Files

The package ships with small BED files under `inst/extdata` and a runnable benchmark script under `inst/benchmarks`:

-   `query_peaks.bed`
-   `subject_genes.bed`
-   `benchmark_overlaps.R`

```{r extdata}
c(
  list.files(system.file("extdata", package = "fastRanges")),
  list.files(system.file("benchmarks", package = "fastRanges"))
)
```

## Session Info

```{r session-info}
sessionInfo()
```
