---
title: "fastRanges: A Practical Introduction to Genomic Interval Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fastRanges: A Practical Introduction to Genomic Interval Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r vignette-options, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

`fastRanges` provides high-throughput interval operations for Bioconductor
workflows built on `IRanges` and `GRanges`.

For genomic data, the central data type is `GRanges`. Each row is one genomic
interval with:

- a sequence name such as `"chr1"`
- a genomic range (`start`, `end`, or `width`)
- a strand (`"+"`, `"-"`, or `"*"`)

Typical analyses ask questions such as:

- Which peaks overlap genes or promoters?
- How many query ranges overlap each subject annotation?
- Which subject is nearest to each query?
- How can the same subject annotation be reused across many batches of query
  intervals?

`fastRanges` is designed to answer those questions while keeping its grammar
close to the Bioconductor overlap APIs in `IRanges` and `GenomicRanges`.

```{r setup}
library(fastRanges)
library(GenomicRanges)
library(S4Vectors)

data("fast_ranges_example", package = "fastRanges")
query <- fast_ranges_example$query
subject <- fast_ranges_example$subject
```

## A Mental Model

The package uses the same two-object language as Bioconductor:

- `query`: the intervals you are asking about
- `subject`: the intervals you compare the query against

If a query peak overlaps three subject genes, then:

- `fast_find_overlaps()` returns three hit pairs
- `fast_count_overlaps()` returns `3` for that query row
- `fast_overlaps_any()` returns `TRUE` for that query row

That distinction is the main decision point when choosing a function.

## Which Function Should You Use?

| Goal | Function | Main output |
|---|---|---|
| Get every matching pair | `fast_find_overlaps()` | `Hits` |
| Count matches per query | `fast_count_overlaps()` | integer vector |
| Ask only yes/no | `fast_overlaps_any()` | logical vector |
| Build a tabular overlap join | `fast_overlap_join()` | `data.frame` |
| Keep only matching query rows | `fast_semi_overlap_join()` | `data.frame` |
| Keep only non-matching query rows | `fast_anti_overlap_join()` | `data.frame` |
| Reuse the same subject many times | `fast_build_index()` | `fast_ranges_index` |
| Find nearest subject | `fast_nearest()` | `DataFrame` |
| Summarize overlap counts by subject metadata | `fast_count_overlaps_by_group()` | matrix |
| Summarize subject scores over overlaps | `fast_overlap_aggregate()` | numeric vector |
| Merge or transform intervals themselves | `fast_reduce()`, `fast_disjoin()`, `fast_gaps()` | ranges |
| Summarize coverage in bins | `fast_tile_coverage()` | `data.frame` |

## The Example Data

The package ships with a small reproducible example:

```{r inspect-example}
query
subject
```

In a real analysis, `query` might be ChIP-seq peaks and `subject` might be
genes, exons, enhancers, promoters, or other genomic annotations.

## Core Overlap Operations

### Return all hit pairs

```{r core-overlaps}
hits <- fast_find_overlaps(query, subject, threads = 2)
hits
```

This result is a `Hits` object. The important parts are:

- `queryHits(hits)`: row indices from `query`
- `subjectHits(hits)`: matching row indices from `subject`

```{r core-overlaps-inspect}
cbind(
  query_id = S4Vectors::queryHits(hits),
  subject_id = S4Vectors::subjectHits(hits)
)
```

### Count matches per query

```{r core-counts}
counts <- fast_count_overlaps(query, subject, threads = 2)
counts
```

There is one value per query row:

- `0` means no subject matched
- `2` means that query matched two subject ranges

### Ask only whether a match exists

```{r core-any}
any_hits <- fast_overlaps_any(query, subject, threads = 2)
any_hits
```

This is the lightest-weight summary when you need only a logical answer.

## Important Arguments

Most overlap-style functions use the same argument grammar:

- `query`
- `subject`
- `max_gap`
- `min_overlap`
- `type`
- `ignore_strand`
- `threads`
- `deterministic`

### `type`

`type` controls what "match" means.

```text
type = "any"
query  :    |------|
subject: |------|
=> any qualifying overlap is a hit

type = "within"
subject: |--------------|
query  :    |------|
=> query must lie inside subject

type = "start"
query  :    |------|
subject:    |------------|
=> start coordinates must match or be within tolerance

type = "end"
query  :       |------|
subject: |------------|
=> end coordinates must match or be within tolerance

type = "equal"
query  :    |------|
subject:    |------|
=> start and end coordinates must match, or be within tolerance
```

### `max_gap`

`max_gap` controls how far apart two ranges may be and still count as a hit.

- `-1` means true overlap is required
- `0` allows Bioconductor's zero-gap tolerance behavior
- positive values allow nearby, non-overlapping ranges to count in the modes
  that support that interpretation

### `min_overlap`

`min_overlap` is the minimum overlap width, in bases.

- `0` is the least strict setting
- larger values require wider shared overlap

### `ignore_strand`

For `GRanges`:

- `FALSE` uses strand-aware Bioconductor semantics
- `TRUE` ignores strand and compares only genomic coordinates

For `IRanges`, strand does not exist, so this argument has no effect.

### `threads` and `deterministic`

- `threads` controls parallel execution
- `deterministic = TRUE` keeps stable ordering of hit pairs
- `deterministic = FALSE` can be slightly faster for large jobs when order is
  unimportant

## Bioconductor Compatibility

`fastRanges` is intended to stay close to Bioconductor overlap semantics.

The easiest way to see that is to compare outputs directly:

```{r compatibility-check}
ref <- GenomicRanges::findOverlaps(query, subject, ignore.strand = FALSE)

identical(
  cbind(S4Vectors::queryHits(hits), S4Vectors::subjectHits(hits)),
  cbind(S4Vectors::queryHits(ref), S4Vectors::subjectHits(ref))
)
```

That compatibility is important because `fastRanges` is designed to fit into
existing `IRanges` / `GRanges` workflows rather than replace those classes.

## Return Types You Will See Most Often

```{r return-types}
ret_classes <- c(
  hits = class(fast_find_overlaps(query, subject, threads = 1))[1],
  count = class(fast_count_overlaps(query, subject, threads = 1))[1],
  any = class(fast_overlaps_any(query, subject, threads = 1))[1],
  nearest = class(fast_nearest(query, subject))[1],
  join = class(fast_overlap_join(query, subject, threads = 1))[1],
  coverage = class(fast_coverage(query))[1]
)

ret_classes
```

Common return shapes:

- `Hits`: pairwise overlap relationships
- integer vector: one summary value per query
- logical vector: one yes/no value per query
- `DataFrame`: nearest-neighbor tables
- `data.frame`: join or coverage summary tables
- `GRanges` / `IRanges`: transformed intervals
- `RleList` or `Rle`: base-resolution coverage

## Reusing the Same Subject with an Index

If you will query the same `subject` repeatedly, build an index once and reuse
it.

```{r index-build}
subject_index <- fast_build_index(subject)
subject_index
```

Now use that index in later overlap calls:

```{r index-use}
hits_from_index <- fast_find_overlaps(query, subject_index, threads = 2)

identical(
  cbind(S4Vectors::queryHits(hits_from_index), S4Vectors::subjectHits(hits_from_index)),
  cbind(S4Vectors::queryHits(hits), S4Vectors::subjectHits(hits))
)
```

When should you build an index?

- yes: many query batches against the same subject
- yes: benchmarking repeated overlap workloads
- no: one small one-off overlap call
- no: when you need subject metadata directly in a join output

## Overlap Joins

The join family turns overlap hits into beginner-friendly tables.

```{r join-inner}
joined_inner <- fast_overlap_join(query, subject, join = "inner", threads = 2)
head(joined_inner)
```

Important columns:

- `query_id`: row index from `query`
- `subject_id`: matching row index from `subject`
- `query_*`: columns copied from `query`
- `subject_*`: columns copied from `subject`

Left join keeps all query rows:

```{r join-left}
joined_left <- fast_left_overlap_join(query, subject, threads = 2)
head(joined_left)
```

Semi and anti joins keep only query rows:

```{r join-semi-anti}
semi_tbl <- fast_semi_overlap_join(query, subject, threads = 2)
anti_tbl <- fast_anti_overlap_join(query, subject, threads = 2)

head(semi_tbl)
head(anti_tbl)
```

## Nearest and Directional Queries

These functions are useful when direct overlap is not the only biological
question.

```{r nearest-queries}
nearest_tbl <- fast_nearest(query, subject)
head(as.data.frame(nearest_tbl))
```

Interpretation:

- `subject_id` is the nearest subject for each matched query
- `distance = 0` means overlap
- positive distance means the nearest subject is separated by a gap

Directional helpers return subject row indices:

```{r directional-queries}
fast_precede(query, subject)
fast_follow(query, subject)
```

## Grouped Overlap Summaries

Many analyses need more than raw hits. They need counts or scores summarized by
subject annotation.

```{r grouped-setup}
set.seed(1)
S4Vectors::mcols(subject)$group <- sample(
  c("promoter", "enhancer"),
  length(subject),
  replace = TRUE
)
S4Vectors::mcols(subject)$score <- seq_len(length(subject))
```

### Count overlaps by group

```{r grouped-counts}
by_group <- fast_count_overlaps_by_group(
  query,
  subject,
  group_col = "group",
  threads = 2
)

by_group
```

Rows are query ranges. Columns are subject groups.

### Aggregate a subject score over overlaps

```{r grouped-aggregate}
sum_score <- fast_overlap_aggregate(
  query,
  subject,
  value_col = "score",
  fun = "sum",
  threads = 2
)

sum_score
```

This pattern is useful when subject intervals carry signal, annotation scores,
or other numeric metadata.

## Self-Overlaps, Clusters, and Sliding Windows

Self-overlaps compare a range set to itself:

```{r self-overlaps}
self_hits <- fast_self_overlaps(query, threads = 2)
self_hits
```

Clusters identify connected components in the self-overlap graph:

```{r overlap-clusters}
clusters <- fast_cluster_overlaps(query, return = "data.frame")
head(clusters)
```

Sliding-window summaries convert interval data into a regular table:

```{r overlap-windows}
window_counts <- fast_window_count_overlaps(
  query,
  subject,
  window_width = 1000L,
  step_width = 500L,
  threads = 2
)

head(window_counts)
```

## Range Algebra

These functions work on the ranges themselves rather than overlap hit pairs.

```{r range-algebra}
query_reduced <- fast_reduce(query)
query_disjoint <- fast_disjoin(query)
query_gaps <- fast_gaps(query)

c(
  original = length(query),
  reduced = length(query_reduced),
  disjoint = length(query_disjoint),
  gaps = length(query_gaps)
)
```

You can also combine two range sets directly:

```{r range-set-ops}
union_ranges <- fast_range_union(query, subject)
intersect_ranges <- fast_range_intersect(query, subject)
setdiff_ranges <- fast_range_setdiff(query, subject)

c(
  union = length(union_ranges),
  intersect = length(intersect_ranges),
  setdiff = length(setdiff_ranges)
)
```

## Coverage and Binned Coverage

Coverage answers: "how many ranges cover each base position?"

```{r coverage}
query_cov <- fast_coverage(query)
query_cov
```

For plotting and downstream summaries, tile-based coverage is often easier to
work with:

```{r tile-coverage}
query_tiles <- fast_tile_coverage(
  query,
  tile_width = 1000L,
  step_width = 1000L
)

head(query_tiles)
```

## Iterating over Large Query Sets

Use the iterator API when you do not want to materialize all hits at once.

```{r iterator}
iter <- fast_find_overlaps_iter(query, subject_index, chunk_size = 2L, threads = 2)

fast_iter_has_next(iter)
chunk1 <- fast_iter_next(iter)
chunk1
fast_iter_has_next(iter)
```

To start over:

```{r iterator-reset}
fast_iter_reset(iter)
all_iter_hits <- fast_iter_collect(iter)
length(all_iter_hits)
```

`fast_iter_collect()` gathers the remaining chunks from the current iterator
position. If you already consumed some chunks and want everything again, reset
the iterator first.

## Saving and Loading an Index

```{r index-io}
tmp_index <- tempfile(fileext = ".rds")

fast_save_index(subject_index, tmp_index)
loaded_index <- fast_load_index(tmp_index)
fast_index_stats(loaded_index)

unlink(tmp_index)
```

This is useful when index construction is expensive and the same subject set
will be queried again in a later session.

## A Small Benchmark Example

The chunk below is only a small demonstration. Publication-scale benchmarking
should use the Quarto workflow in `inst/benchmarks`.

```{r benchmark-demo}
base_time <- system.time({
  h_base <- GenomicRanges::findOverlaps(query, subject, ignore.strand = FALSE)
})

fast_time <- system.time({
  h_fast <- fast_find_overlaps(query, subject_index, ignore_strand = FALSE, threads = 2)
})

rbind(
  genomic_ranges = base_time[1:3],
  fast_ranges = fast_time[1:3]
)

c(base_hits = length(h_base), fast_hits = length(h_fast))
```

For larger benchmark studies, see:

- `inst/benchmarks/README.md`
- `inst/benchmarks/benchmark_bioc.qmd`

## Package Example Files

The package also ships example BED files and benchmarking helpers:

```{r extdata}
c(
  list.files(system.file("extdata", package = "fastRanges")),
  list.files(system.file("benchmarks", package = "fastRanges"))
)
```

## Session Info

```{r session-info}
sessionInfo()
```
