---
title: "fastRanges Reviewer Response Assessment"
subtitle: "Post hoc interpretation of reviewer_response_check.R outputs"
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    code-summary: "Show code"
execute:
  echo: true
  warning: false
  message: false
params:
  results_dir: "reviewer_response_results"
  figures_dir: "reviewer_response_figures"
---

## Goal

This report assesses the saved outputs produced by
`inst/benchmarks/04/reviewer_response_check.R`. It is intended to answer the
reviewer's concerns systematically using the recorded compatibility checks,
timing summaries, preprocessing costs, and peak RAM measurements.

Unlike the benchmark script itself, this document does not rerun the reviewer
workload. It reads the saved CSV outputs, regenerates interpretation figures,
and summarizes what the run demonstrates and what it does not.

## Scope

The reviewer raised five practical concerns:

1. semantic differences from `findOverlaps()`
2. default single-thread performance
3. the effect of `deterministic`
4. preprocessing/index construction cost
5. memory usage

The sections below follow that structure.

## Setup

```{r}
required_pkgs <- c("ggplot2", "dplyr", "tidyr", "scales")
missing_pkgs <- required_pkgs[!vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing_pkgs) > 0) {
  stop("Missing required packages: ", paste(missing_pkgs, collapse = ", "))
}

library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)

input_dir <- dirname(knitr::current_input())

resolve_path <- function(path, mustWork = FALSE) {
  candidates <- c(
    path,
    file.path(input_dir, path),
    file.path(getwd(), path)
  )
  for (candidate in unique(candidates)) {
    if (file.exists(candidate) || dir.exists(candidate)) {
      return(normalizePath(candidate, winslash = "/", mustWork = mustWork))
    }
  }
  normalizePath(file.path(input_dir, path), winslash = "/", mustWork = mustWork)
}

results_dir <- resolve_path(params$results_dir, mustWork = TRUE)
figures_dir <- resolve_path(params$figures_dir, mustWork = FALSE)
if (!dir.exists(figures_dir)) {
  dir.create(figures_dir, recursive = TRUE)
}

read_tbl <- function(name) {
  path <- file.path(results_dir, paste0(name, ".csv"))
  if (!file.exists(path)) {
    stop("Missing result table: ", path)
  }
  utils::read.csv(path, stringsAsFactors = FALSE, check.names = FALSE)
}

save_plot <- function(plot, filename, width = 10, height = 6, dpi = 320) {
  path <- file.path(figures_dir, filename)
  ggplot2::ggsave(path, plot = plot, width = width, height = height, dpi = dpi)
  path
}

theme_pub <- function() {
  theme_bw(base_size = 13) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title.position = "plot",
      legend.position = "bottom",
      strip.background = element_rect(fill = "grey95"),
      axis.text.x = element_text(color = "black"),
      axis.text.y = element_text(color = "black")
    )
}

label_engine <- function(x) {
  x <- gsub("^GenomicRanges_findOverlaps$", "GenomicRanges::findOverlaps", x)
  x <- gsub("^GenomicRanges_GNCList$", "GenomicRanges::findOverlaps(GNCList)", x)
  x <- gsub("^fastRanges_default$", "fast_find_overlaps() default", x)
  x <- gsub("^fastRanges_build_index$", "fast_build_index()", x)
  x <- gsub("^GenomicRanges_GNCList_build$", "GNCList()", x)
  x <- gsub("^fastRanges_direct", "fastRanges direct", x)
  x <- gsub("^fastRanges_indexed", "fastRanges indexed", x)
  x <- gsub("_t", " | t=", x)
  x <- gsub("_det_", " | deterministic=", x)
  x <- gsub("_false", "FALSE", x)
  x <- gsub("_true", "TRUE", x)
  x
}

parse_engine <- function(x) {
  data.frame(
    engine = x,
    mode = dplyr::case_when(
      grepl("^fastRanges_default$", x) ~ "default",
      grepl("^fastRanges_direct", x) ~ "direct",
      grepl("^fastRanges_indexed", x) ~ "indexed",
      grepl("^GenomicRanges_findOverlaps$", x) ~ "findOverlaps",
      grepl("^GenomicRanges_GNCList$", x) ~ "GNCList",
      TRUE ~ "other"
    ),
    threads = dplyr::case_when(
      grepl("_t1_", x) ~ 1L,
      grepl("_t8_", x) ~ 8L,
      grepl("_t96_", x) ~ 96L,
      TRUE ~ NA_integer_
    ),
    deterministic = dplyr::case_when(
      grepl("_det_true$", x) ~ "TRUE",
      grepl("_det_false$", x) ~ "FALSE",
      grepl("^fastRanges_default$", x) ~ "TRUE (default)",
      TRUE ~ NA_character_
    ),
    stringsAsFactors = FALSE
  )
}

run_meta <- read_tbl("run_metadata")
pkg_versions <- read_tbl("package_versions")
semantic <- read_tbl("semantic_checks")
pre_raw <- read_tbl("preprocessing_raw")
pre_summary <- read_tbl("preprocessing_summary")
workload <- read_tbl("reviewer_workload")
perf_raw <- read_tbl("performance_raw")
perf_summary <- read_tbl("performance_summary")
det_summary <- read_tbl("determinism_thread_summary")
peak_ram <- read_tbl("peak_ram_summary")

perf_aug <- perf_summary %>% left_join(parse_engine(perf_summary$engine), by = "engine")
det_aug <- det_summary %>% left_join(parse_engine(det_summary$engine), by = "engine")
peak_aug <- peak_ram %>% left_join(parse_engine(peak_ram$engine), by = "engine")

semantic <- semantic %>%
  mutate(
    concern_label = c(
      empty_ranges = "Empty ranges",
      circular_sequences = "Circular sequences",
      GRangesList_support = "GRangesList",
      select_argument = "select="
    )[concern],
    status_label = case_when(
      status %in% c("match", "supported") ~ "Acceptable",
      status %in% c("unsupported_or_error", "missing") ~ "Explicit gap",
      TRUE ~ "Mismatch"
    )
  )

best_fast <- perf_summary %>%
  filter(grepl("^fastRanges_", engine)) %>%
  arrange(median_sec) %>%
  slice(1)

default_fast <- perf_summary %>% filter(engine == "fastRanges_default")
baseline_gr <- perf_summary %>% filter(engine == "GenomicRanges_findOverlaps")
baseline_gnc <- perf_summary %>% filter(engine == "GenomicRanges_GNCList")
```

## Run Context

```{r}
run_meta
```

```{r}
pkg_versions
```

The recorded reviewer workload used:

```{r}
workload
```

This corresponds to:

- `query_n = 3,000,000`
- `subject_n = 1,000,000`
- mean query width of about `255`
- mean subject width of about `22,011`

## Executive Assessment

```{r}
summary_table <- tibble::tibble(
  question = c(
    "Does this run show semantic parity with GenomicRanges?",
    "Is default fastRanges behavior faster than findOverlaps()?",
    "Is best-throughput fastRanges faster than findOverlaps()?",
    "Is best-throughput fastRanges faster than GNCList()?",
    "Is index construction expensive?",
    "Does deterministic ordering materially affect throughput?"
  ),
  assessment = c(
    "No. The saved run still reports circular mismatch and missing select=.",
    sprintf("No. Default fastRanges median was %.3f sec vs %.3f sec for findOverlaps().",
            default_fast$median_sec, baseline_gr$median_sec),
    sprintf("Yes. Best fastRanges median was %.3f sec, a %.2fx speedup vs findOverlaps().",
            best_fast$median_sec, best_fast$speedup_vs_baseline),
    sprintf("Yes. Best fastRanges median was %.3f sec vs %.3f sec for GNCList().",
            best_fast$median_sec, baseline_gnc$median_sec),
    sprintf("No. fast_build_index() median was %.3f sec vs %.3f sec for GNCList() construction.",
            pre_summary$median_sec[pre_summary$engine == "fastRanges_build_index"],
            pre_summary$median_sec[pre_summary$engine == "GenomicRanges_GNCList_build"]),
    "Yes. deterministic=TRUE largely removes the multithreading benefit in this run."
  )
)

summary_table
```

The important constraint is that this result snapshot reflects the package state
captured when `reviewer_response_check.R` was run. It is not automatically the
same as the current branch state.

## 1. Semantic Compatibility

```{r}
semantic %>%
  select(concern_label, scenario, status, reference_supported, fastRanges_supported, message)
```

### Interpretation

- Empty ranges matched the Bioconductor reference in this run.
- Circular sequences did not match the reference in this run.
- `GRangesList` was not supported in this run.
- `select=` was missing in this run.

That means this result set represents a pre-fix compatibility snapshot rather
than a fully updated reviewer-response state.

```{r}
p_semantic <- ggplot(semantic, aes(x = concern_label, y = 1, fill = status_label)) +
  geom_tile(color = "white", linewidth = 0.7) +
  geom_text(aes(label = status), size = 4) +
  scale_fill_manual(
    values = c(
      "Acceptable" = "#2A9D8F",
      "Explicit gap" = "#E9C46A",
      "Mismatch" = "#D1495B"
    ),
    drop = FALSE
  ) +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(
    title = "Semantic status recorded by reviewer_response_check.R",
    subtitle = "This reflects the package state at the time of the captured run",
    x = NULL,
    fill = "Assessment"
  ) +
  theme_pub()

p_semantic
save_plot(p_semantic, "semantic_status_heatmap.png", width = 10, height = 3.8)
```

## 2. Preprocessing Cost

```{r}
pre_summary
```

The preprocessing result is straightforward:

- `fast_build_index()` was faster than constructing `GNCList()`
- the `fastRanges` index object was also smaller in this run

```{r}
p_pre_time <- ggplot(
  pre_summary %>% mutate(engine_label = label_engine(engine)),
  aes(x = reorder(engine_label, median_sec), y = median_sec, fill = engine_label)
) +
  geom_col(width = 0.72, show.legend = FALSE) +
  geom_errorbar(aes(ymin = q25_sec, ymax = q75_sec), width = 0.14, linewidth = 0.6) +
  coord_flip() +
  scale_y_continuous(labels = label_number(accuracy = 0.001)) +
  labs(
    title = "Preprocessing runtime",
    subtitle = "Median build time across iterations",
    x = NULL,
    y = "Median runtime (sec)"
  ) +
  theme_pub()

p_pre_time
save_plot(p_pre_time, "preprocessing_runtime.png", width = 8, height = 4.8)
```

```{r}
p_pre_mem <- ggplot(
  pre_raw %>% mutate(engine_label = label_engine(engine)),
  aes(x = reorder(engine_label, object_size_mb), y = object_size_mb, fill = engine_label)
) +
  geom_col(width = 0.72, show.legend = FALSE) +
  coord_flip() +
  scale_y_continuous(labels = label_number(accuracy = 0.1)) +
  labs(
    title = "Preprocessing object size",
    subtitle = "Resident object size of the built index structure",
    x = NULL,
    y = "Object size (MB)"
  ) +
  theme_pub()

p_pre_mem
save_plot(p_pre_mem, "preprocessing_object_size.png", width = 8, height = 4.8)
```

## 3. Overall Performance on the Reviewer Workload

```{r}
perf_summary %>%
  mutate(engine_label = label_engine(engine)) %>%
  select(engine_label, median_sec, q25_sec, q75_sec, speedup_vs_baseline)
```

Two different conclusions come out of this table, and they should not be mixed:

1. Default `fast_find_overlaps()` was slower than `GenomicRanges::findOverlaps()`.
2. Best-throughput `fastRanges` settings were substantially faster than both
   `findOverlaps()` and `findOverlaps(GNCList(...))`.

```{r}
p_perf_all <- ggplot(
  perf_summary %>% mutate(engine_label = label_engine(engine)),
  aes(x = reorder(engine_label, median_sec), y = median_sec, fill = engine_label)
) +
  geom_col(width = 0.72, show.legend = FALSE) +
  geom_errorbar(aes(ymin = q25_sec, ymax = q75_sec), width = 0.15, linewidth = 0.6) +
  coord_flip() +
  scale_y_continuous(labels = label_number(accuracy = 0.1)) +
  labs(
    title = "Runtime on the reviewer's synthetic workload",
    subtitle = "Lower is better; bars show median runtime and interquartile range",
    x = NULL,
    y = "Median runtime (sec)"
  ) +
  theme_pub()

p_perf_all
save_plot(p_perf_all, "reviewer_workload_runtime_all.png", width = 10, height = 7.2)
```

```{r}
p_speedup <- ggplot(
  perf_summary %>%
    filter(engine != "GenomicRanges_findOverlaps") %>%
    mutate(engine_label = label_engine(engine)),
  aes(x = reorder(engine_label, speedup_vs_baseline), y = speedup_vs_baseline, fill = engine_label)
) +
  geom_col(width = 0.72, show.legend = FALSE) +
  geom_hline(yintercept = 1, linetype = 2, color = "grey40") +
  coord_flip() +
  scale_y_continuous(labels = label_number(accuracy = 0.01)) +
  labs(
    title = "Speedup relative to GenomicRanges::findOverlaps()",
    subtitle = "Values above 1 indicate faster runtime than the baseline",
    x = NULL,
    y = "Speedup (x)"
  ) +
  theme_pub()

p_speedup
save_plot(p_speedup, "reviewer_workload_speedup_vs_findOverlaps.png", width = 10, height = 6.5)
```

### Interpretation

The workload does not support a blanket statement that `fastRanges` is always
faster. It supports a narrower, more accurate statement:

- default single-thread deterministic behavior was slower than
  `GenomicRanges::findOverlaps()`
- high-throughput `fastRanges` settings were substantially faster
- the best recorded configuration was `fastRanges indexed`, `threads = 96`,
  `deterministic = FALSE`

## 4. Deterministic Ordering and Thread Count

This is the most important performance nuance in the result set.

```{r}
det_aug %>%
  filter(mode %in% c("direct", "indexed")) %>%
  arrange(mode, deterministic, threads) %>%
  select(engine, mode, threads, deterministic, median_sec, speedup_vs_baseline)
```

```{r}
p_det_runtime <- ggplot(
  det_aug %>% filter(mode %in% c("direct", "indexed")),
  aes(x = threads, y = median_sec, color = mode, linetype = deterministic)
) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.2) +
  scale_x_continuous(breaks = c(1, 8, 96)) +
  scale_color_manual(
    values = c("direct" = "#D1495B", "indexed" = "#00798C"),
    labels = c("direct" = "fastRanges direct", "indexed" = "fastRanges indexed")
  ) +
  labs(
    title = "Runtime across thread counts and deterministic settings",
    subtitle = "The deterministic sort dominates runtime in this captured run",
    x = "Threads",
    y = "Median runtime (sec)",
    color = "Mode",
    linetype = "deterministic"
  ) +
  theme_pub()

p_det_runtime
save_plot(p_det_runtime, "deterministic_thread_runtime.png", width = 10, height = 6)
```

```{r}
speedup_by_mode <- det_aug %>%
  filter(mode %in% c("direct", "indexed")) %>%
  group_by(mode, deterministic) %>%
  mutate(speedup_vs_1_thread = median_sec[threads == 1L] / median_sec) %>%
  ungroup()

p_det_scaling <- ggplot(
  speedup_by_mode,
  aes(x = threads, y = speedup_vs_1_thread, color = mode, linetype = deterministic)
) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.2) +
  geom_hline(yintercept = 1, linetype = 2, color = "grey40") +
  scale_x_continuous(breaks = c(1, 8, 96)) +
  scale_color_manual(
    values = c("direct" = "#D1495B", "indexed" = "#00798C"),
    labels = c("direct" = "fastRanges direct", "indexed" = "fastRanges indexed")
  ) +
  labs(
    title = "Thread scaling relative to 1 thread",
    subtitle = "Substantial scaling is observed only when deterministic = FALSE",
    x = "Threads",
    y = "Speedup vs 1 thread (x)",
    color = "Mode",
    linetype = "deterministic"
  ) +
  theme_pub()

p_det_scaling
save_plot(p_det_scaling, "deterministic_thread_scaling.png", width = 10, height = 6)
```

### Interpretation

This run strongly supports the reviewer's observation that `deterministic`
matters.

- With `deterministic = TRUE`, runtimes remained in the `~68-82 sec` range even
  at higher thread counts.
- With `deterministic = FALSE`, runtime dropped to about `10.4-10.7 sec` at 96
  threads.

So the defensible statement is not merely that `fastRanges` supports
multithreading. It is that the package's strongest throughput mode requires
`deterministic = FALSE`.

## 5. Peak RAM

```{r}
peak_aug %>%
  mutate(engine_label = label_engine(engine)) %>%
  select(engine_label, peak_ram_mb, elapsed_sec)
```

```{r}
p_peak_ram <- ggplot(
  peak_aug %>% mutate(engine_label = label_engine(engine)),
  aes(x = reorder(engine_label, peak_ram_mb), y = peak_ram_mb, fill = engine_label)
) +
  geom_col(width = 0.72, show.legend = FALSE) +
  coord_flip() +
  scale_y_continuous(labels = label_number(accuracy = 1)) +
  labs(
    title = "Peak RAM by engine configuration",
    subtitle = "Reported only because peakRAM was available for this run",
    x = NULL,
    y = "Peak RAM (MiB)"
  ) +
  theme_pub()

p_peak_ram
save_plot(p_peak_ram, "peak_ram_by_engine.png", width = 10, height = 5.8)
```

### Interpretation

This captured run does not support a simple claim that `fastRanges` always uses
substantially more memory than the Bioconductor baselines.

- `fastRanges_default` was the heaviest configuration in this run
- best-throughput `fastRanges` settings were close to the Bioconductor baselines
- all non-default configurations clustered in a relatively narrow range around
  `~3.94-3.97 GiB`

That means the memory story is more nuanced than "fastRanges uses 50% more
memory". At minimum it depends on configuration.

## 6. Reviewer-Facing Conclusions

The most defensible conclusions from this saved result set are:

1. The captured run still reflects unresolved compatibility gaps at the time it
   was executed.
   - `select=` was absent
   - circular overlap behavior did not match the reference
   - `GRangesList` was unsupported

2. The reviewer's default-performance concern is supported by this run.
   - default `fast_find_overlaps()` was slower than `findOverlaps()`

3. The reviewer's broader performance claim is not supported once explicit
   throughput settings are benchmarked.
   - `fastRanges indexed`, `threads = 96`, `deterministic = FALSE` achieved a
     median runtime of about `10.4 sec`
   - this was about `3.45x` faster than `GenomicRanges::findOverlaps()`
   - and faster than `findOverlaps(GNCList(...))`

4. The reviewer's deterministic caveat is strongly supported.
   - high throughput depends on `deterministic = FALSE`

5. Preprocessing cost is not a weakness in this run.
   - `fast_build_index()` was faster than `GNCList()` construction and produced
     a smaller object

## Bottom Line

This output set supports a narrow but strong reviewer-response narrative:

- the package state captured here was not yet fully compatible with
  `findOverlaps()` semantics
- however, under explicit high-throughput settings, `fastRanges` was clearly
  faster than the Bioconductor baselines on the reviewer's own synthetic
  workload
- the main performance caveat is not lack of threading, but the cost of
  deterministic ordering

That is the right interpretation to carry forward into a formal reviewer
response. It is also a good reason to regenerate this assessment after the
semantic fixes on the current branch, so that the compatibility table reflects
the updated implementation rather than this older snapshot.

